% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      


My PhD was far from a solitary journey. I would like to acknowledge the direct and indirect contributions of the wonderful people I met and worked with over the last four years. My heartfelt gratitude goes to Dr. Carl Anderson, my PhD supervisor. His guidance, encouragement and thoughtful feedback over the past four years have made this intellectual journey truly life-changing.\\ %He constantly encouraged me to not only think critically and rigorously about my analyses, but to also think about how one's work should contribute to a wider scientific discourse. I also thank him for giving me the freedom to pursue scientific questions that I found interesting, and for helping me develop a sense of intellectual ownership of my work.\\

I would also like to extend my gratitude to my friends and colleagues in the Anderson laboratory and in the Department of Human Genetics at the Sanger Institute. In particular, I am grateful to Dr. Nikolaos I Panousis whom I worked with on the MacroMap project and who gave me the opportunity to lead the alternative splicing arm of the project. Special thanks go out to Dr. Laura Fachal who often discussed with me many of the analyses described here and provided invaluable feedback. I had the privilige of working with other members of the laboratory whose expertise on different genetic analyses was extremely helpful: Dr. Aleksejs Sazonovs, Dr. Qian Zhang, Dr Marcus Tutert and every other member of the Anderson laboratory. I am indebted to every participant and patient who donated samples to the HipSci project, the UKB IBD Genetic Cosortium, the IBD BioResource and the UK BioBank.\\

Finally, I am incredibly grateful to my family, my ultimate motivation throughout this challenging journey. This journey would not have been possible without the love and support of my parents: Mohamed El Gawany and Samia El Saidy. My parents have been, and will always be, my ultimate mentors. Their love, sacrifice and wisdom have guided me through the years. I will always be grateful to my wife Sondos Khedr. Her unceasing support of me has kept me going and her impressive strength and resilience have inspired me at every difficult moment. I owe you so much. I was lucky to be supported by a group of amazing friends who were always there for me: Mohamed Yehia Kamel, Ahmed Abdelbaky, Samuel Gabra, Hani Saleh, Youmna Hussein, Samer Abdelmoeti, John Danial and Abdelhamid Yousef. My heartfelt gratitude goes out to each and every one of you. 


\end{acknowledgements}
